---
author: Sebastian Barfort
title: "Social Data Science"
subtitle: "Exercise 5.4.1"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  beamer_presentation:
    keep_tex: no
    theme: metropolis
    latex_engine: xelatex
    slide_level: 2
    incremental: no
    fig_width: 7
    fig_height: 6
    fig_caption: false
    includes:
      in_header: header.tex
fontsize: 10pt
classoption: compress
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
hook_output = knitr::knit_hooks$get('output')
knitr::knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
knitr::opts_chunk$set(
              dev= "pdf",
               fig.width=4.25,
               fig.height=2.5,
               fig.show="hold",
               fig.lp="fig:",
               fig.align = "center",
               dpi = 300,
               cache=TRUE,
               par=TRUE,
               echo=TRUE,
               message=FALSE,
               warning=FALSE)
```

\LARGE Exercise 5.4.1 

## Read data

### `intrade`

```{r}
library("readr")
gh.link = "https://raw.githubusercontent.com/"
user.repo = "kosukeimai/qss/"
branch = "master/"
link = "PREDICTION/intrade08.csv"
data.link = paste0(gh.link, user.repo, branch, link)
intrade.08 = read_csv(data.link)
```

```{r}
link = "PREDICTION/intrade12.csv"
data.link = paste0(gh.link, user.repo, branch, link)
intrade.12 = read_csv(data.link)
```

##

 Name                Description
 ------------------- ---------------------------------------------------------
 `day`               Date of the session 
 `statename`         Full name of each state 
 `state`             Abbreviation of each state 
 `PriceD`            Predicted vote share of D Nominee's market
 `PriceR`            Predicted vote share of R Nominee's market 
 `VolumeD`           Total session trades of D Nominee's market
 `VolumeR`           Total session trades of R Nominee's market 

## 

### `pres`

```{r}
link = "PREDICTION/pres08.csv"
data.link = paste0(gh.link, user.repo, branch, link)
pres.08 = read_csv(data.link)
```

```{r}
link = "PREDICTION/pres12.csv"
data.link = paste0(gh.link, user.repo, branch, link)
pres.12 = read_csv(data.link)
```

##

 Name                 Description
 -------------------- ---------------------------------------------------------
 `state.name`         Full name of state (only in `pres2008`)
 `state`              Two letter state abbreviation
 `Obama`              Vote percentage for Obama
 `McCain`             Vote percentage for McCain
 `EV`                 Number of electoral college votes for this state

## 

### `polls`

```{r}
link = "PREDICTION/polls08.csv"
data.link = paste0(gh.link, user.repo, branch, link)
polls.08 = read_csv(data.link)
```

```{r}
link = "PREDICTION/polls12.csv"
data.link = paste0(gh.link, user.repo, branch, link)
polls.12 = read_csv(data.link)
```

## 

 Name                 Description
 -------------------- ---------------------------------------------------------
 `state`              Abbreviated name of state in which poll was conducted 
 `Obama`              Predicted support for Obama (percentage)
 `Romney`             Predicted support for Romney (percentage) 
 `Pollster`           Name of organization conducting poll 
 `middate`            Middle of the period when poll was conducted 

## 5.4.1.1

```{r}
library("dplyr")
df.08 = intrade.08 %>% 
  left_join(pres.08)
```

## 

```{r}
df.08 = df.08 %>% 
  mutate( # gen variables
    days.to.election = as.Date("2008-11-04") - 
      as.Date(day),
    obama.intmarg = PriceD - PriceR,
    obama.actmarg = Obama - McCain
  ) %>% 
  mutate( # gen sign indicator
    obama.intmarg.sign = sign(obama.intmarg),
    obama.actmarg.sign = sign(obama.actmarg)
  )
```

## 

```{r}
df.08 %>% 
  filter(days.to.election == 1) %>%
  filter(obama.intmarg.sign != 
           obama.actmarg.sign) %>% 
  count()
```

## 

```{r, echo = FALSE}
library("ggplot2")
p = ggplot(df.08, 
           aes(x = days.to.election, 
               y = EV))
p = p + geom_point() + 
  geom_hline(yintercept = 370, 
             color = "red", 
             linetype = "dashed")
p
```

